
---
title: "Bellabeat Case Study: Data Analysis & Business Insights"
author: "Dona Manojlovic"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
    number_sections: true
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(gridExtra)
```
```{r load-data, echo=TRUE}
load("merged_data.RData")
```

# Introduction

Bellabeat, a wellness technology company, aims to unlock insights from smart device usage data to guide marketing strategy for its mobile app. This analysis uses Fitbit data to explore trends in activity, sleep, and heart rate.

# 1. Activity Overview

## 1.1 Summary Statistics
```{r}
summary_activity <- merged_data %>%
  summarise(
    avg_steps = round(mean(total_steps, na.rm = TRUE), 0),
    max_steps = max(total_steps, na.rm = TRUE),
    min_steps = min(total_steps, na.rm = TRUE),
    avg_distance = round(mean(total_distance, na.rm = TRUE), 2),
    avg_calories = round(mean(calories, na.rm = TRUE), 0),
    max_calories = max(calories, na.rm = TRUE),
    min_calories = min(calories, na.rm = TRUE)
  )
knitr::kable(summary_activity, caption = "Summary of User Activity")
```

## 1.2 Average Steps by Weekday
```{r}
merged_data <- merged_data %>% mutate(weekday = weekdays(date))
p1 <- merged_data %>%
  group_by(weekday) %>%
  summarise(avg_steps = mean(total_steps, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(weekday, -avg_steps), y = avg_steps)) +
  geom_bar(stat = "identity", fill = "#69b3a2") +
  labs(title = "Average Steps by Weekday", x = "Weekday", y = "Steps") +
  theme_minimal()
p1
```

## 1.3 Steps vs Calories Burned
```{r}
p2 <- merged_data %>%
  filter(!is.na(total_steps), !is.na(calories)) %>%
  ggplot(aes(x = total_steps, y = calories)) +
  geom_point(alpha = 0.5, color = "#1f78b4") +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(title = "Steps vs Calories Burned", x = "Total Steps", y = "Calories") +
  theme_minimal()
p2
```

# 2. Sleep Patterns

## 2.1 Sleep Summary
```{r}
summary_sleep <- merged_data %>%
  summarise(
    avg_minutes_asleep = mean(total_minutes_asleep, na.rm = TRUE),
    avg_time_in_bed = mean(total_time_in_bed, na.rm = TRUE),
    min_sleep = min(total_minutes_asleep, na.rm = TRUE),
    max_sleep = max(total_minutes_asleep, na.rm = TRUE)
  )
knitr::kable(summary_sleep, caption = "Summary of Sleep Duration")
```

## 2.2 Sleep Duration Histogram
```{r}
p3 <- merged_data %>%
  filter(!is.na(total_minutes_asleep), total_minutes_asleep > 0) %>%
  ggplot(aes(x = total_minutes_asleep)) +
  geom_histogram(binwidth = 30, fill = "#2c7fb8", color = "white") +
  labs(title = "Distribution of Sleep Duration", x = "Minutes Asleep", y = "Count") +
  theme_minimal()
p3
```

# 3. Heart Rate Trends

## 3.1 Summary Heart Rate by Day
```{r}
summary_hr <- merged_data %>%
  group_by(date) %>%
  summarise(avg_hr = mean(avg_hr, na.rm = TRUE)) %>%
  head(10)
knitr::kable(summary_hr, caption = "Sample of Average Heart Rate Over Time")
```

## 3.2 Heart Rate Over Time
```{r}
p4 <- merged_data %>%
  group_by(date) %>%
  summarise(avg_hr = mean(avg_hr, na.rm = TRUE)) %>%
  ggplot(aes(x = date, y = avg_hr)) +
  geom_line(color = "#fc8d62", linewidth = 1) +
  labs(title = "Average Daily Heart Rate", x = "Date", y = "Heart Rate (bpm)") +
  theme_minimal()
p4
```

## 3.3 Steps vs Heart Rate
```{r}
p5 <- merged_data %>%
  filter(!is.na(total_steps), !is.na(avg_hr)) %>%
  ggplot(aes(x = total_steps, y = avg_hr)) +
  geom_point(alpha = 0.5, color = "#1f78b4") +
  geom_smooth(method = "lm", se = FALSE, color = "#e34a33") +
  labs(title = "Steps vs Average Heart Rate", x = "Total Steps", y = "Avg HR (bpm)") +
  theme_minimal()
p5
```

# 4. Business Recommendations

- Encourage users to set and meet step goals, especially on weekends
- Provide smart sleep insights based on consistency
- Use heart rate patterns to personalize workout suggestions
- Introduce weekly summary insights to motivate users

# 5. Conclusion

This analysis provides Bellabeat with valuable insights into user health patterns. By implementing data-informed features into their app, Bellabeat can enhance user engagement and promote healthier habits.
